Statistical Relational Models as Polyadic Measure Algebras

### 2020-03-06 paper.tex
The start of an example

\documentclass{article}
\usepackage[utf8]{inputenc}
\usepackage[UKenglish]{babel}
\usepackage[UKenglish]{isodate}
\usepackage{fullpage}
\usepackage{amsthm}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{mathtools}
\usepackage[capitalise]{cleveref}
\usepackage{bm}

@@

\Crefname{property}{Property}{Properties}
\DeclareMathOperator{\WMC}{WMC}
\DeclareMathOperator{\nWMC}{nWMC}
\DeclareMathOperator{\id}{id}

@@

\begin{definition}[\cite{gaifman1964concerning}] \label{def:measure}
  A \emph{(strictly positive) measure} on a Boolean algebra $\mathbf{B}$ is a
  function $m : \mathbf{B} \to [0, 1]$ such that:
  \begin{enumerate}
  \item $m(1) = 1$, and $m(x) > 0$ for $x \ne
    0$; \label[property]{property:values}
  \item $m(x \lor y) = m(x) + m(y)$ for all $x, y \in \mathbf{B}$ whenever $x
    \land y = 0$. \label[property]{property:additive}
  \end{enumerate}
\end{definition}

@@

\begin{definition}
  Let $\mathbf{B}$ be a finite Boolean algebra, let $L$ be a subset of
  $\mathbf{B}$ such that every atom $m$ can be uniquely expressed as $m =
  \bigwedge_{i \in I} l_i$ for some $\{ l_i \}_{i \in I} \subseteq L$, and let
  $w : L \to \mathbb{R}_{>0}$ be arbitrary. The \emph{weighted model count}
  $\WMC : \mathbf{B} \to \mathbb{R}_{\ge 0}$ is defined as
  \[
    \WMC(a) = \begin{cases}
      0 & \text{if } a = 0 \\
      \prod_{i \in I} w(l_i) & \text{if } M \ni a = \bigwedge_{i \in I} l_i \\
      \sum_{i \in I} \WMC(m_i) & \text{if } \mathbf{B} \setminus (M \cup \{ 0
      \}) \ni a = \bigvee_{i \in I} m_i
    \end{cases}
  \]
  for any $a \in \mathbf{B}$. Furthermore, we define the \emph{normalised
    weighted model count} $\nWMC : \mathbf{B} \to [0, 1]$ as $\nWMC(a) =
  \frac{\WMC(a)}{\WMC(1)}$ for all $a \in \mathbf{B}$.
\end{definition}

@@

% Defining S
Let $T$ be the semigroup of all $I \to I$ transformations. For any $\tau \in
T$, let $\tau_* : X^I \to X^I$ be defined by
\[
  (\tau_* x)_i = x_{\tau i}
\]
for all $x \in X^I$ and $i \in I$. We can then define $\mathbf{S}$ to be a map
from $T$ to Boolean endomorphisms of $\mathbf{A^*}$ defined by
\[
  \mathbf{S}(\tau)p(x) = p(\tau_* x)
\]
for all $x \in X^I$ and $p \in \mathbf{A^*}$.

@@

\begin{definition}
  Similarly to $\bm\exists$, a \emph{constant} $c$ is a mapping from
  $\mathcal{P}(I)$ to Boolean endomorphisms of $\mathbf{A}$ such that:
  \begin{itemize}
  \item $c(\emptyset) = \id_{\mathbf{A}}$;
  \item $c(J \cup K) = c(J)c(K)$;
  \item $c(J)\bm\exists(K) = \bm\exists(K)c(J \setminus K)$;
  \item $\bm\exists(J)c(K) = c(K)\bm\exists(J \setminus K)$;
  \item $c(J)\mathbf{S}(\tau) = \mathbf{S}(\tau)c(\tau^{-1}J)$
  \end{itemize}
  for all $J, K \in \mathcal{P}(I)$ and $\tau \in T$. If $J$ is a singleton
  set $\{ i \}$, we will simply write $c(i)$ instead of $c(J)$.
\end{definition}

@@

% TODO: how to interpret first-order sentences in this algebra
% TODO: define WMC: connect my definition of WMC with the original one
% TODO: condition for a measurable BA to be representable as WMC
% TODO: can I show that a generalisation of WMC is the only possible probability measure (when it exists)?
% TODO: abstraction refinements as homomorphisms.
% TODO: definition of a measure-preserving homomorphism from Thomas Jech
% TODO: a Boolean algebra is approximable if its Stone space is approximable

>>>>

\documentclass{article}
\usepackage[utf8]{inputenc}
\usepackage[UKenglish]{babel}
\usepackage[UKenglish]{isodate}
\usepackage{fullpage}
\usepackage{amsthm}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{mathtools}
\usepackage[capitalise]{cleveref}
\usepackage{bm}
\usepackage{booktabs}

@@

\Crefname{property}{Property}{Properties}
\DeclareMathOperator{\WMC}{WMC}
\DeclareMathOperator{\nWMC}{nWMC}
\DeclareMathOperator{\id}{id}
\DeclareMathOperator{\End}{End}

@@

% TODO: remove the requirement for being strictly positive
\begin{definition}[\cite{gaifman1964concerning}] \label{def:measure}
  A \emph{(strictly positive) measure} on a Boolean algebra $\mathbf{B}$ is a
  function $m : \mathbf{B} \to [0, 1]$ such that:
  \begin{enumerate}
  \item $m(1) = 1$, and $m(x) > 0$ for $x \ne
    0$; \label[property]{property:values}
  \item $m(x \lor y) = m(x) + m(y)$ for all $x, y \in \mathbf{B}$ whenever $x
    \land y = 0$. \label[property]{property:additive}
  \end{enumerate}
\end{definition}

@@

% TODO: weight can be zero
\begin{definition}
  Let $\mathbf{B}$ be a finite Boolean algebra, let $L$ be a subset of
  $\mathbf{B}$ such that every atom $m$ can be uniquely expressed as $m =
  \bigwedge_{i \in I} l_i$ for some $\{ l_i \}_{i \in I} \subseteq L$, and let
  $w : L \to \mathbb{R}_{>0}$ be arbitrary. The \emph{weighted model count}
  $\WMC : \mathbf{B} \to \mathbb{R}_{\ge 0}$ is defined as
  \[
    \WMC(a) = \begin{cases}
      0 & \text{if } a = 0 \\
      \prod_{i \in I} w(l_i) & \text{if } M \ni a = \bigwedge_{i \in I} l_i \\
      \sum_{i \in I} \WMC(m_i) & \text{if } \mathbf{B} \setminus (M \cup \{ 0
      \}) \ni a = \bigvee_{i \in I} m_i
    \end{cases}
  \]
  for any $a \in \mathbf{B}$. Furthermore, we define the \emph{normalised
    weighted model count} $\nWMC : \mathbf{B} \to [0, 1]$ as $\nWMC(a) =
  \frac{\WMC(a)}{\WMC(1)}$ for all $a \in \mathbf{B}$.
\end{definition}

@@

% Defining S
Let $T$ be the semigroup of all $I \to I$ transformations. For any $\tau \in
T$, let $\tau_* : X^I \to X^I$ be defined by
\[
  (\tau_* x)_i = x_{\tau i}
\]
for all $x \in X^I$ and $i \in I$. For any (Boolean/polyadic) algebra
$\mathbf{C}$, let $\End(\mathbf{C})$ denote the set of all its endomorphisms. We
can then define $\mathbf{S}$ to be a map $\mathbf{S} : T \to \End(\mathbf{A^*})$
defined by
\[
  \mathbf{S}(\tau)p(x) = p(\tau_* x)
\]
for all $x \in X^I$ and $p \in \mathbf{A^*}$.

@@

\begin{definition}
  Similarly to $\bm\exists$, a \emph{constant} $c$ is a map $c: \mathcal{P}(I)
  \to \End(\mathbf{A})$ (Boolean endomorphisms?) such that:
  \begin{itemize}
  \item $c(\emptyset) = \id_{\mathbf{A}}$;
  \item $c(J \cup K) = c(J)c(K)$;
  \item $c(J)\bm\exists(K) = \bm\exists(K)c(J \setminus K)$;
  \item $\bm\exists(J)c(K) = c(K)\bm\exists(J \setminus K)$;
  \item $c(J)\mathbf{S}(\tau) = \mathbf{S}(\tau)c(\tau^{-1}J)$
  \end{itemize}
  for all $J, K \in \mathcal{P}(I)$ and $\tau \in T$. If $J$ is a singleton
  set $\{ i \}$, we will simply write $c(i)$ instead of $c(J)$.
\end{definition}

@@

\section{From First-Order Logic to Polyadic Algebras}

\begin{definition}[\cite{givant2008introduction}] \label{def:boolean_ideal}
  An \emph{ideal} in a Boolean algebra $\mathbf{B}$ is a subset $M \subseteq
  \mathbf{B}$ such that:
  \begin{itemize}
  \item $0 \in M$;
  \item $x \lor y \in M$ for all $x, y \in M$;
  \item $x \land y \in M$ for all $x \in M$ and $y \in \mathbf{B}$.
  \end{itemize}
  For any subset $S \subseteq \mathbf{B}$, the \emph{ideal generated by $S$} is
  the smallest ideal $M$ such that $S \subseteq M$.
\end{definition}

Note that \cref{def:boolean_ideal} gives us a simple characterisation of an
ideal generated by a set of atoms.

\begin{lemma}
  Let $\mathbf{B}$ be a Boolean algebra, and let $S \subseteq \mathbf{B}$ be a
  set of atoms. The ideal $I$ generated by $S$ is defined by the following:
  \begin{itemize}
  \item $0 \in I$,
  \item $S \subseteq I$,
  \item $x \lor y \in I$ for all $x, y \in I$.
  \end{itemize}
\end{lemma}

\begin{definition}[\cite{givant2008introduction}]
  Let $\mathbf{B}$ be a Boolean algebra, and let $I$ be an ideal in
  $\mathbf{B}$. The \emph{quotient algebra} $\mathbf{B}/I$ is a Boolean algebra
  on equivalence classes of elements of $\mathbf{B}$ (with operations defined
  pointwise) based on the equivalence relation
  \[
    x \sim y \quad \iff \quad x + y \in I
  \]
  where $x + y = (x \land \neg y) \lor (y \land \neg x)$ is the symmetric
  difference operation (written as a sum because it can interpreted as the
  `additive' part of the corresponding Boolean ring).
\end{definition}

In order to make the example algebras easily describable, our example programs
will have to be tiny. Consider the following ProbLog
\cite{DBLP:conf/ijcai/RaedtKT07} program:
\begin{align*}
  1.0 &\dblcolon \mathsf{p}(a, b).\\
  0.5 &\dblcolon \mathsf{p}(X, X) \coloneq \mathsf{p}(X, Y);\, \mathsf{p}(Y, X).
\end{align*}
Let $L = \{ \mathsf{p}(a, a), \mathsf{p}(a, b), \mathsf{p}(b, a), \mathsf{p}(b,
b) \}$ be the set of all possible ground atoms. Let $\mathbf{B}$ be the
Boolean algebra freely generated by $L$ (see, e.g.,
\cite{givant2008introduction} for more on free Boolean algebras). Then
$\mathbf{B}$ will have sixteen atoms ranging from $\mathsf{p}(a, a) \land
\mathsf{p}(a, b) \land \mathsf{p}(b, a) \land \mathsf{p}(b, b)$ to
$\neg\mathsf{p}(a, a) \land \neg\mathsf{p}(a, b) \land \neg\mathsf{p}(b, a)
\land \neg\mathsf{p}(b, b)$. The weight function $w : L \to \mathbb{R}_{\ge 0}$
defined by
\[
  w(l) = \begin{cases}
    1 & \text{if } l = \mathsf{p}(a, b) \\
    0.5 & \text{if } l \in \{ \mathsf{p}(a, a), \mathsf{p}(b, b) \} \\
    0 & \text{if } l = \mathsf{p}(b, a) \\
    1-w(l') & \text{if } l = \neg l'
  \end{cases}
\]
for all $l \in L$ defines a WMC measure over $\mathbf{B}$. Note that while we
could define an ideal generated by $\{ \mathsf{p}(b, a), \neg\mathsf{p}(a, b)
\}$ and take the quotient of $\mathbf{B}$ by that ideal to get a Boolean algebra
with a strictly positive measure, this would put zero-probability queries
outside of our algebras, i.e., we would not be able to ask a question whose
answer is zero.

\begin{table}
  \centering
  \caption{Example elements of $\mathbf{A}$ as maps $X^I \to \mathbf{B}$, with
    $a : \mathcal{P}(I) \to \End(\mathbf{A})$ as one of two possible constants.}
  \label{tbl:examples}
  \begin{tabular}{ll}
    \toprule
    Element of $\mathbf{A}$ & Action on $X^I$ \\
    \midrule
    $p = \mathbf{S}(\id)p = \bm\exists(\emptyset)p = a(\emptyset)p = b(\emptyset)p$ & $(x_1, x_2) \mapsto \mathsf{p}(x_1, x_2)$ \\
    $\bm\exists(1)p$ & $(x_1, x_2) \mapsto \mathsf{p}(a, x_2) \lor \mathsf{p}(b, x_2)$ \\
    $\bm\exists(2)p$ & $(x_1, x_2) \mapsto \mathsf{p}(x_1, a) \lor \mathsf{p}(x_1, b)$ \\
    $\bm\exists(I)p$ & $(x_1, x_2) \mapsto \mathsf{p}(a, a) \lor \mathsf{p}(a, b) \lor \mathsf{p}(b, a) \lor \mathsf{p}(b, b)$ \\
    $\mathbf{S}(\{ 1 \mapsto 1, 2 \mapsto 1 \})p$ & $(x_1, x_2) \mapsto \mathsf{p}(x_1, x_1)$ \\
    $\mathbf{S}(\{ 1 \mapsto 2, 2 \mapsto 1 \})p$ & $(x_1, x_2) \mapsto \mathsf{p}(x_2, x_1)$ \\
    $\mathbf{S}(\{ 1 \mapsto 2, 2 \mapsto 2 \})p$ & $(x_1, x_2) \mapsto \mathsf{p}(x_2, x_2)$ \\
    $a(1)p$ & $(x_1, x_2) \mapsto \mathsf{p}(a, x_2)$ \\
    $a(2)p$ & $(x_1, x_2) \mapsto \mathsf{p}(x_1, a)$ \\
    $a(I)p$ & $(x_1, x_2) \mapsto \mathsf{p}(a, a)$ \\
    \bottomrule
  \end{tabular}
\end{table}

Finally, let $\mathbf{A}$ be the functional polyadic algebra $X^I \to
\mathbf{B}$ for $I = \{1, 2\}$ and $X = \{ a, b \}$\footnote{$X$ cannot (or
  should not) have constants that do not occur in $\mathbf{B}$.}. The elements
of $X^I$ can then be represented as tuples $(x_1, x_2)$ for some $x_1, x_2 \in
X$. See \cref{tbl:examples} for example elements of $\mathbf{A}$ which consists
of a single predicate function $p$ and operators $\bm\exists, \mathbf{S}, a, b,
\neg, \land, \lor$, the last three of which are defined pointwise.

% TODO: 1-2 examples of how a probability is calculated
% TODO: condition for a measurable BA to be representable as WMC
% TODO: can I show that a generalisation of WMC is the only possible probability measure (when it exists)?
% TODO: abstraction refinements as homomorphisms.
% TODO: definition of a measure-preserving homomorphism from Thomas Jech
% TODO: a Boolean algebra is approximable if its Stone space is approximable
% TODO: maybe the measure on BA should not be normalised?
% TODO: how do measures transfer to quotient algebras? Naturally!
% TODO: extension to infinite atomic Boolean algebras

\section{Homomorphisms}

\begin{definition}[\cite{givant2008introduction}]
  Let $\mathbf{A}$ and $\mathbf{B}$ be Boolean algebras. A \emph{Boolean
    homomorphism} from $\mathbf{A}$ to $\mathbf{B}$ is a map $f : \mathbf{A} \to
  \mathbf{B}$ such that:
  \begin{itemize}
  \item $f(x \land y) = f(x) \land f(y)$,
  \item $f(x \lor y) = f(x) \lor f(y)$,
  \item $f(\neg x) = \neg f(x)$
  \end{itemize}
  for all $x, y \in \mathbf{A}$.
\end{definition}

\begin{definition}[\cite{halmos2016algebraic}]
  Given two polyadic algebras $\mathbf{A}$ and $\mathbf{B}$, a \emph{polyadic
    homomorphism} from $\mathbf{A}$ to $\mathbf{B}$ is a Boolean homomorphism $f
  : \mathbf{A} \to \mathbf{B}$ such that
  \begin{itemize}
  \item $f\mathbf{S}(\tau)p = \mathbf{S}(\tau)fp$,
  \item $f\bm\exists(J)p = \bm\exists(J)fp$
  \end{itemize}
  for all $\tau \in T$, $p \in \mathbf{A}$, and $J \subseteq I$.
\end{definition}
